---
title: "Dealing with label switching: relabelling in Bayesian mixtures by pivotal units"
author: "Leonardo Egidi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ref.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette we explore the relabelling pivotal method proposed in @egidi2018relabelling through the `pivmet` package. First of all, we load the package:

```{r load}
library(pivmet)
```


There are two main functions designed for our procedure: 

- `piv_MCMC`: 

- performs MCMC sampling for Gaussian mixture models launching the `JAGSrun` function of the `bayesmix` package for univariate models or the `run.jags` function of the `runjags` package for bivariate models; - post-processes the chains and randomly swithes their values; 
- builds a co-association matrix $C$ for the $N$ statistical units, where the single cell $c_{ip}$ is the fraction of times the unit $i$ and the unit $p$ belong to the same group along the $H$ MCMC iterations;
- extracts some pivotal units via the internal function `piv_sel`, one for each of the pre-specified $k$ groups.

- `piv_rel`: performs the relabelling algorithm using the $k$ pivotal units as groups identifiers. It yields the relabelled chains and the corresponding posterior estimates.

## Example: bivariate data

Suppose $\boldsymbol{y}_i \sim \sum_{j=1}^{k}\pi_{k}\mathcal{N}_{2}(\boldsymbol{\mu}_{k}, \boldsymbol{\Sigma})$. We may generate Gaussian mixture data through the function `piv_sim`, specifying the desired number of groups $k$. The argument `W` handles the weights for a nested mixture, in which each $j$-th component is in turn modelled as a two-component mixture.


```{r nested}
N  <- 2000
k  <- 3
nMC <- 2000
M1 <- c(-45,8)
M2 <- c(45,.1)
M3 <- c(100,8)
Mu <- matrix(rbind(M1,M2,M3),c(k,2))
stdev    <- cbind(rep(1,k), rep(200,k))
Sigma.p1 <- matrix(c(stdev[1,1],0,0,stdev[1,1]),
nrow=2, ncol=2)
Sigma.p2 <- matrix(c(stdev[1,2],0,0,stdev[1,2]),
 nrow=2, ncol=2)
W   <- c(0.2,0.8)
sim <- piv_sim(N = N, k = k, Mu = Mu,
  stdev = stdev, Sigma.p1 = Sigma.p1, Sigma.p2 = Sigma.p2, W = W)
plot(sim$y, xlab="y[,1]", ylab="y[,2]")
```

The function ```piv_MCMC``` requires only three mandatory arguments: data ```y```, number of component ```k``` and number of MCMC iterations ```nMC```. It performs JAGS sampling using the ```bayesmix``` package for univariate data and ```runjags``` package for bivariate data. After $H$ MCMC iterations, the function implements a clustering procedure on $k$ groups---through the optional argument `clustering` the user may choose among agglomerative or divisive hierarchical clustering---. Using the latent formulation for mixture models, we denote with $[Z_i]_h$ the group allocation of the $i$-th unit at the $h$-th iteration. The function builds a co-association matrix $C$ across the MCMC sample with generic element:

$$c_{ip} = \frac{n_{ip}}{H}=\frac{1}{H} \sum_{h=1}^{H}|[Z_i]_h=[Z_p]_h|,$$

where $|\cdot|$ denotes the event indicator and $n_{ip}$ is the number of times the units $i, \ p$ belong to the same group along the sampling. Using this matrix, we may extract some units, one for each group, which
are (pairwise) separated with (posterior) probability one (that
is, the posterior probability of any two of them being in the
same group is zero). We call them *pivots*, and denote them with: $i_1,\ldots,i_k$. With the optional argument `piv.criterion`, the user may choose among three (plus one) procedures for extracting the pivotal units. For group $j$ containing $J_j$ units, one can choose:

-  $i^{*}$ that maximizes $\sum_{p \in J_j}c_{ip}$ (`maxsumint`);

- $i^{*}$ that maximizes $\sum_{p \in J_j}c_{ip}- \sum_{p \not\in J_j}c_{ip}$ (`maxsumdiff`);

- $i^{*}$ that minimizes $\sum_{p \not\in J_j}c_{ip}$ (`maxsumnoint`).

Alternatively, when $k <5$, the user has the optional pivotal criterion `MUS` [@egidi2018mus] based on a sequential search within the matrix $C$.



```{r mcmc, message =  FALSE, warning = FALSE}
res <- piv_MCMC(y = sim$y, k= k, nMC =nMC)
```

Once we obtain posterior estimates, label switching is likely to occurr. For such a reason, we need to relabel our chains, and the pivotal units previously detected play a central role, yielding to the following relabelling:

\begin{align*} 
[\mu_{j}]_h=&[\mu_{Z_{i_{j}}}]_h \\
[Z_{i}]_h =j & \mbox{ for } i : [Z_i]_h=[Z_{i_{j}}]_h.\\
\end{align*}

The function `piv_rel` performs the procedure above, and for such task it only needs the `mcmc = rel` argument and `nMC`. Once we correctly relabel the chains, we may plot the relabelled outputs through the function `piv_plot`, with different options for the argument `type`:

- `chains`: plot the relabelled chains;
- `estimates`: plot the point estimates for the parameters of interest;
- `estimates_hist`: plot the point estimates against the histogram of the data;
- `iter`: plot the proportions of valid MCMC iterations.


```{r pivotal_rel, fig.show='hold', fig.align='center'}
rel <- piv_rel(mcmc=res, nMC = nMC)
piv_plot(y = y, mcmc = res, rel_est = rel, type = "chains")
piv_plot(y = y, mcmc = res, rel_est = rel, type = "estimates")
piv_plot(y = y, mcmc = res, rel_est = rel, type = "estimates_hist")
```


## References

